Grow with the flow: a latitudinal cline in physiology is associated with more variable precipitation in \textit{Mimulus cardinalis}

### 2016-10-26 ms/ms.Rnw
New figure of climate variable procedure

%lost on page 11 - extracted and used climate variables
%cartoon of whole procedure for climate variables
%figure showing correlation b/t bio12 and 
%state that VSURF importance is different scales in different models
%moved up three hypotheses part to introduction
%play up lack of GxE and ???
%** correct # of Pop x Treatment
% model of growth chamber
% need to explain 'importance'

@@

%* put in biomass allocation cline
% TO DO:
% 1. Is water regime on figure 1 correct? (data is in office)
% 2. Double check spatially averaged analysis
% 3. For Figure_PC1vLat: plot axes could be range maps with areas highlighted (x-axis) and short to tall plants (y-axis)
% 4. Do anything with biomass data? Currently, NO.

@@

We found that a population's position along TraitPC1 correlated strongly with the latitude or origin (see Results) and next used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M. cardinalis}. Hypothesis 1: if a single climatic factor explained more trait variation than latitude alone, this would suggest that that factor is a selective agent underlying the latitudinal cline in our 16 focal populations. Hypothesis 2: if multiple climatic factors together explained trait variation,  we interpreted this as evidence that multiple climatic factors together have generated the latitudinal cline. We hereafter refer to factors identified in this analysis as `Climate-TraitPC1' variables. In addition, to help eliminate potentially spurious correlations between TraitPC1 and climate, we tested for overlap between climatic variables that best predict latitude of all \textit{M. cardinalis} occurrence records, not just the 16 focial populations. We refer to these climatic factors as `Climate-Latitude' variables. The logic is that climatic factors associated with both TraitPC1 and latitude for all populations are more likely to be important selective agents than climatic factors that happen to correlate with TraitPC1 but do not covary with latitude throughout the \textit{M. cardinalis} range. We selected Climate-Latitude and Climate-TraitPC1 variables independently with Variable Selection Using Random Forest (VSURF) algorithm in the R package \pkg{VSURF} version 1.0.3 \citep{Genuer_etal_2016}. VSURF ranks variables by their importance over regression trees in the forest. We kept only variables selected for prediction, the most stringent criterion.

% This is now stated above.
%If latitude is strongly correlated with one or two climatic variables that are responsible for the cline, then there should be one or two climatic variables that are identified in both the Climate-Latitude and Climate-TraitPC1 analyses \ala{and the correlation magnitudes should be high? maybe this language is not technically correct (variable importance more accurate?) but the point is that this prediction needs something about count and the strength}. If latitude is correlated with several climatic agents of selection that are individually weak, but add up to a strong latitudinal cline, then several climatic variables should be weakly correlated \ala{fix this to mirror prior}.

@@

@@

>>>>

%lost on page 11 - extracted and used climate variables
%cartoon of whole procedure for climate variables
% model of growth chamber
% need to explain 'importance'

@@

%* put in biomass allocation cline
% TO DO:
% 2. Double check spatially averaged analysis
% 3. For Figure_PC1vLat: plot axes could be range maps with areas highlighted (x-axis) and short to tall plants (y-axis)
% 4. Do anything with biomass data? Currently, NO.

@@

We found that a population's position along TraitPC1 correlated strongly with the latitude or origin (see Results) and next used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M. cardinalis}. Hypothesis 1: if a single climatic factor explained more trait variation than latitude alone, this would suggest that that factor is a selective agent underlying the latitudinal cline in our 16 focal populations. Hypothesis 2: if multiple climatic factors together explained trait variation,  we interpreted this as evidence that multiple climatic factors together have generated the latitudinal cline. We hereafter refer to factors identified in this analysis as `Climate-TraitPC1' variables. In addition, to help eliminate potentially spurious correlations between TraitPC1 and climate, we tested for overlap between climatic variables that best predict latitude of all \textit{M. cardinalis} occurrence records, not just the 16 focial populations. We refer to these climatic factors as `Climate-Latitude' variables. The logic is that climatic factors associated with both TraitPC1 and latitude for all populations are more likely to be important selective agents than climatic factors that happen to correlate with TraitPC1 but do not covary with latitude throughout the \textit{M. cardinalis} range. We selected Climate-Latitude and Climate-TraitPC1 variables independently with Variable Selection Using Random Forest (VSURF) algorithm in the R package \pkg{VSURF} version 1.0.3 \citep{Genuer_etal_2016}. VSURF ranks variables by their importance over regression trees in the forest. We kept only variables selected for prediction, the most stringent criterion. A visual overview of how we selected climatic variables is depicted in Fig~\ref{fig:Fig_VarSelect}.

@@

<<Fig_VarSelect, echo=FALSE, results=hide, warning=FALSE, message=FALSE>>=

  library(raster)
	library(rgeos)
  library(extrafont)

  # Get maps of California and Oregon
  usa <- getData('GADM', country = "USA", level = 1, path = pathDat)
  states <- gSimplify(usa[usa@data$NAME_1 %in% c("California", "Oregon",
                                                 "Nevada", "Idaho"), ], 
                      tol = 0.01, topologyPreserve = TRUE)

  # Load thinned occurences used in Climate-Lat analyses and convert to Spatial class
  occ <- read.csv(paste(pathDat, "/thinned1OccLatLon.csv", sep = ""), 
                  row.names = NULL)
  coordinates(occ) <- ~Longitude+Latitude
  projection(occ) <- CRS('+proj=longlat');

  # Convert focal population coordinates to Spatial class
  foc <- focClim[, c("Longitude", "Latitude")]
  coordinates(foc) <- ~Longitude+Latitude
  projection(foc) <- CRS('+proj=longlat');

  # Clip and map
  clipExtent <- as(extent(floor(min(occ$Longitude) * 10) / 10, 
	   											ceiling(max(occ$Longitude) * 10) / 10, 
	  											floor(min(occ$Latitude) * 10) / 10, 
		  										ceiling(max(occ$Latitude) * 10) / 10), 
			  					 "SpatialPolygons")
  proj4string(clipExtent) <- CRS(proj4string(states))
  states <- gIntersection(states, clipExtent, byid = TRUE)

  # Transform to Albert's Equal Area Projection
  projAEA <- sprintf("+proj=aea +lat_1=%s +lat_2=%s +lat_0=%s +lon_0=%s +x_0=0 +y_0=0",
                     extent(clipExtent)[3], extent(clipExtent)[4], 
		  							 extent(clipExtent)[1], extent(clipExtent)[2])
  states <- spTransform(states, CRS(projAEA))
  occAEA <- spTransform(occ, CRS(projAEA))
  focAEA <- spTransform(foc, CRS(projAEA))

  # PLOT
  pdf(paste(pathFig, "/Figure_VarSelectBase.pdf", sep = ""), w = 6, h = 7, 
	  	useDingbat = F)
  loadfonts()
  mat <- matrix(c(1, 4, 7,
	  							1, 4, 7,
		  						1, 4, 7,
			  					2, 5, 8,
				  				2, 5, 8,
					  			3, 6, 9,
						  		3, 6, 9), ncol = 3, byrow = T)
  layout(mat)

  # A. Example Climate-Latitude variable

	  ## Panel 1: Map
	  par(mai = rep(0.5, 4), family = "sans")
	  plot(states, main = "A. Climate-Latitude variables")
	  points(occAEA, pch = 19, cex = 0.5)

	  # Position of Cuyamaca Rancho population in device units
	  xCUR <- grconvertX(focAEA$Longitude["CUR"], from = "user", to = "device")
	  yCUR <- grconvertY(focAEA$Latitude["CUR"], from = "user", to = "device")

	  ## Panel 2: Cartoon of point versus spatial average climate estimate
	  par(mai = rep(0, 4), family = "LM Roman 10")

	    ### River 1
	    x1 <- seq(-1, 1, length.out = 1e2)
	    r1 <- function(x) exp(x) - exp(0) + sin(x * 10) / 10
	    y1 <- r1(x1)

	    ### River 2
	    x2 <- seq(0, 1, length.out = 1e2)
	    r2 <- function(x) exp(-x) - exp(0) + sin(x * 10) / 10
	    y2 <- r2(x2)

	  plot(0, 0, type = "n", xlim = c(-1, 1), ylim = c(-1, 1), axes = F)
	  clip(-1, 1, -1, 1)
	  points(x1, y1, type = "l", col = "blue", lwd = 5)
	  points(x2, y2, type = "l", col = "blue", lwd = 5)

	  # Draw buffer area
	  x <- seq(-1, 0, length.out = 1e3)
	  clip(-2, 2, -2, 2)
	  polygon(c(x, rev(x)), c(sqrt(1 - x ^ 2), rep(1.1, 1e3)), 
	          col = "white", border = "white", lwd = 2)
	  polygon(c(x + 1, rev(x + 1)), c(sqrt(1 - (x + 1) ^ 2), rep(1.1, 1e3)), 
	          col = "white", border = "white", lwd = 2)
	  polygon(c(x, rev(x)), -c(sqrt(1 - x ^ 2), rep(1.1, 1e3)), 
	          col = "white", border = "white", lwd = 2)
	  polygon(c(x + 1, rev(x + 1)), -c(sqrt(1 - (x + 1) ^ 2), rep(1.1, 1e3)), 
	          col = "white", border = "white", lwd = 2)

	  # Draw point and label
	  points(0, 0, pch = 19, cex = 3)
	  text(0, 0, labels = expression(italic(x[paste(i, ",", 0)])), adj = 1.5, 
	       cex = 1.5)
	  # lines(c(grconvertX(xCUR, from = "device", to = "user"), 0),
	  #     c(grconvertY(yCUR, from = "device", to = "user"), 0),
	  #       lwd = 2)
    text(0, 1, labels = expression(Climate[italic(i)] == 
           italic(x[paste(i, ",", 0)])), cex = 1.5, pos = 3, offset = 1)

	  ## Panel 3: Climate-Latitude relationship
	  set.seed(01152016)
  	y <- (occ$Latitude - mean(occ$Latitude)) / sd(occ$Latitude) + 
		      rnorm(length(occ), 0, sqrt(0.1))
	  par(mai = c(0.6, 0.6, 1/25, 1/25), family = "sans")
	  plot(y, occ$Latitude, las = 1, axes = F, frame.plot = T, 
		  	 xlim = c(-2.5, 2.5), ylim = c(32, 45), cex.lab = 1.5,
			   pch = 19, col = "grey", xlab = "", ylab = "Latitude of origin")
	  title(xlab = "Climate variable", cex.lab = 1.5, line = 1)

  	for (i in seq(32, 44, 2)) {
		  axis(2, at = i, labels = bquote(.(i)*degree), las = 1)
	  	}
	  fit <- lm(occ$Latitude ~ y)
	  lines(range(y), range(predict(fit)), lwd = 2)

   # B. Example Climate-Trait variable, point estimated

  	## Panel 1: Map
	  par(mai = rep(0.5, 4), family = "sans")
	  plot(states, main = "B. Climate-Trait variables")
	  mtext(text = "point estimated", side = 3)
	  points(focAEA, pch = 19, cex = 0.5)

	  # Position of Cuyamaca Rancho population in device units
	  xCUR <- grconvertX(focAEA$Longitude["CUR"], from = "user", to = "device")
	  yCUR <- grconvertY(focAEA$Latitude["CUR"], from = "user", to = "device")

	  ## Panel 2: Cartoon of point versus spatial average climate estimate
	  par(mai = rep(0, 4), family = "LM Roman 10")
	  plot(0, 0, type = "n", xlim = c(-1, 1), ylim = c(-1, 1), axes = F)
	  clip(-1, 1, -1, 1)
	  points(x1, y1, type = "l", col = "blue", lwd = 5)
	  points(x2, y2, type = "l", col = "blue", lwd = 5)

	  # Draw buffer area
	  x <- seq(-1, 0, length.out = 1e3)
	  clip(-2, 2, -2, 2)
	  polygon(c(x, rev(x)), c(sqrt(1 - x ^ 2), rep(1.1, 1e3)), 
	          col = "white", border = "white", lwd = 2)
	  polygon(c(x + 1, rev(x + 1)), c(sqrt(1 - (x + 1) ^ 2), rep(1.1, 1e3)), 
	          col = "white", border = "white", lwd = 2)
	  polygon(c(x, rev(x)), -c(sqrt(1 - x ^ 2), rep(1.1, 1e3)), 
	          col = "white", border = "white", lwd = 2)
	  polygon(c(x + 1, rev(x + 1)), -c(sqrt(1 - (x + 1) ^ 2), rep(1.1, 1e3)), 
	          col = "white", border = "white", lwd = 2)

	  # Draw point and label
	  points(0, 0, pch = 19, cex = 3)
	  text(0, 0, labels = expression(italic(x[paste(i, ",", 0)])), adj = 1.5, 
	       cex = 1.5)
	  # lines(c(grconvertX(xCUR, from = "device", to = "user"), 0),
	  #       c(grconvertY(yCUR, from = "device", to = "user"), 0),
	  #       lwd = 2)
    text(0, 1, labels = expression(Climate[italic(i)] == 
           italic(x[paste(i, ",", 0)])), cex = 1.5, pos = 3, offset = 1)

	  ## Panel 3: Climate-trait relationship
	  par(mai = c(0.6, 0.6, 1/25, 1/25), family = "sans")
	  y <- (foc$Latitude - mean(foc$Latitude)) / sd(foc$Latitude) + 
		      rnorm(length(foc), 0, sqrt(0.1))
	  plot(y, foc$Latitude, las = 1, axes = F, frame.plot = T, 
		  	 xlim = c(-2.5, 2.5), ylim = c(32, 45),
			   pch = 19, col = "grey", xlab = "", ylab = "")
	  title(xlab = "Climate variable", ylab = "Trait", cex.lab = 1.5, line = 1)
	  fit <- lm(foc$Latitude ~ y)
	  lines(range(y), range(predict(fit)), lwd = 2)

  # C. Example Climate-Trait variable, spatially averaged

  	## Panel 1: Map
  	par(mai = rep(0.5, 4), family = "sans")
  	plot(states, main = "C. Climate-Trait variables")
  	tmp <- raster(states)
  	mtext(text = "spatially averaged", side = 3)
  	points(focAEA, pch = 21, cex = 2, col = "black", bg = "white")
	  points(focAEA, pch = 19, cex = 0.5)

	  # Position of Cuyamaca Rancho population in device units
	  xCUR <- grconvertX(focAEA$Longitude["CUR"], from = "user", to = "inches")
	  yCUR <- grconvertY(focAEA$Latitude["CUR"], from = "user", to = "inches")

  	## Panel 2: Cartoon of point versus spatial average climate estimate
	  par(mai = rep(0, 4), family = "LM Roman 10")

  		### Climate grid

  			#### Functions to find closest point along "stream"
				fn <- function(par, x, y) {
					if (x < 0) {
						d1 <- sqrt((par - x) ^ 2 + (r1(par) - y) ^ 2)
						return(d1)
					} else {
						d1 <- sqrt((par - x) ^ 2 + (r1(par) - y) ^ 2)
						d2 <- sqrt((par - x) ^ 2 + (r2(par) - y) ^ 2)
						return(min(c(d1, d2)))
					}
				}

				getStreamDist <- function(x, y) {
					fit <- optim(x, fn, x = x, y = y, method = "Brent", lower = -2, 
											 upper = 2)
					retval <- setNames(c(fit$par, fit$value), c("s", "dist"))
					return(retval)
				}

			  #### "Climate" is fxn of distance from stream and pos along x-axis
			  xy <- expand.grid(x = seq(-0.95, 0.95, 0.1), 
			                    y = seq(-0.95, 0.95, 0.1))
			  xyz <- cbind(xy, 
			               t(apply(xy, 1, function(X) getStreamDist(X[1], X[2]))))
			  xyz$z <- xyz$dist + xyz$x * 0.3
			  xyz$z <- xyz$z - min(xyz$z)
			  xyz$z <- xyz$z / max(xyz$z)

		  plot(0, 0, type = "n", xlim = c(-1, 1), ylim = c(-1, 1), axes = F)
		  clip(-1, 1, -1, 1)
		  points(x1, y1, type = "l", col = "blue", lwd = 5)
		  points(x2, y2, type = "l", col = "blue", lwd = 5)

		  apply(xyz, 1, function(X) rect(X["x"] - 0.05, X["y"] - 0.05, 
																	   X["x"] + 0.05, X["y"] + 0.05,
														 			   col = rgb(0, 0, 1, alpha = X["z"]), 
																	   border = NA))

			# Draw buffer area and circle
			x <- seq(-1, 0, length.out = 1e3)
			clip(-2, 2, -2, 2)
			polygon(c(x, rev(x)), c(sqrt(1 - x ^ 2), rep(1.1, 1e3)), 
			        col = "white", border = "white", lwd = 2)
			polygon(c(x + 1, rev(x + 1)), c(sqrt(1 - (x + 1) ^ 2), rep(1.1, 1e3)), 
			        col = "white", border = "white", lwd = 2)
			polygon(c(x, rev(x)), -c(sqrt(1 - x ^ 2), rep(1.1, 1e3)), 
			        col = "white", border = "white", lwd = 2)
			polygon(c(x + 1, rev(x + 1)), -c(sqrt(1 - (x + 1) ^ 2), rep(1.1, 1e3)), 
			        col = "white", border = "white", lwd = 2)

			x <- seq(-1, 1, length.out = 1e3)
			polygon(c(x, rev(x)), c(sqrt(1 - x ^ 2), -sqrt(1 - x ^ 2)),
			        border = "black", lwd = 2)

			# Draw points and label
	    points(0, 0, pch = 19, cex = 3)
	    text(0, 0, labels = expression(italic(x[paste(i, ",", 0)])), adj = 1.5, 
	         cex = 1.5)
	    # Start line from outside of point
	    # point radius = cex * size of point = cex * 0.75 * character height
	    # ptRad <- 0.1 * 2 * 0.75 * par()$cin[2]
	    # lines(c(grconvertX(xCUR - ptRad, from = "inches", to = "user"), -1),
	    #     c(grconvertY(yCUR, from = "inches", to = "user"), 0),
	    #       lwd = 2)
      text(0, 1, labels = expression(Climate[italic(i)] == 
           sum(over(x[paste(italic(i), ",", italic(j))], italic(n)), 
               scriptstyle(italic(j) == 1), italic(n))), 
           cex = 1.5, pos = 3, offset = 1)

      points(c(-2/3, -1/3, 1/2, 1/3, 2/3),
             c(r1(-2/3), r1(-1/3), r1(1/2), r2(1/3), r2(2/3)),
             cex = 2, pch = 21, bg = "grey", col = "black")

      text(-2/3 + 0.15, r1(-2/3) - 0.15, cex = 1.5,
           labels = expression(italic(x[paste(i, ",", 1)])))
      text(-1/3, r1(-1/3) - 0.15, cex = 1.5,
           labels = expression(italic(x[paste(i, ",", 2)])))
      text(1/3 - 0.15, r2(1/3) - 0.15, cex = 1.5,
           labels = expression(italic(x[paste(i, ",", 3)])))
      text(1/2 - 0.15, r1(1/2) + 0.15, cex = 1.5,
           labels = expression(italic(x[paste(i, ",", 4)])))
      text(2/3, r2(2/3) + 0.15, cex = 1.5,
           labels = expression(italic(x[paste(i, ",", 5)])))

      par(family = "sans")
      str <- "62-km\nbuffer"
      rect(-0.4 - strwidth(str, cex = 1.75) / 2,
           0.4 - strheight(str, cex = 1.75) / 2,
           -0.4 + strwidth(str, cex = 1.75) / 2,
           0.4 + strheight(str, cex = 1.75) / 2, 
           col = rgb(1, 1, 1, alpha = 0.5), border = NA)
      text(-0.4, 0.4, labels = str, cex = 1.5)

      ## Panel 3: Climate-trait relationship

		  par(mai = c(0.6, 0.6, 1/25, 1/25), family = "sans")
		  plot(y, foc$Latitude, las = 1, axes = F, frame.plot = T, 
				   xlim = c(-2.5, 2.5), ylim = c(32, 45),
				   pch = 19, col = "grey", xlab = "", ylab = "")
		  title(xlab = "Climate variable", ylab = "Trait", cex.lab = 1.5, 
		        line = 1)
		  lines(range(y), range(predict(fit)), lwd = 2)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure of variable selection algorithm
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[h!]
	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_VarSelectBase.pdf}}
	\usefont{T1}{phv}{m}{n}
	\fontsize{10}{12}
	\selectfont
	\caption[Climatic variable selection]{Overview of process for identifying putative climatic selective drivers underlying latitudinal cline. We looked for climate variables that explained both the latitude of \Sexpr{2 * nrow(thinnedOccClimate)} \textit{M. cardinalis} occurences (`Climate-Latitude variables') and with traits (`Climate-Trait variables'). (A) For Climate-Latitude variables we extracted climate data from the specific location where a population was collected ($x_{i,0}$) throughout California and Oregon, USA (shown in map). For Climate-Trait variables, we extracted climatic data in two ways, but only for the 16 focal populations. (B) For point-estimated Climate-Trait variables, we extracted climate only from where a population was collected ($x_{i,0}$). (C) For spatially averaged Climate-Trait variables, we averaged over 1000 points in a 62-km radius climatic neighborhood ($x_{i,1}, x_{i,2}, \dots$), but only along stream habitats as \textit{M. cardinalis} is riparian.}
	\label{fig:Fig_VarSelect}
\end{figure}

@@

  	# lines between panels
  	# clip(grconvertX(0, from = "ndc", to = "user"),
  	#     grconvertX(1, from = "ndc", to = "user"),
  	#     grconvertY(0.05, from = "ndc", to = "user"),
  	#     grconvertY(0.95, from = "ndc", to = "user"))
	  # abline(v = grconvertX(1/3, from = "ndc", to = "user"))

### 2016-10-26 ms/ms.tex
New figure of climate variable procedure

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure of variable selection algorithm
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[h!]
	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_VarSelectBase.pdf}}
	\usefont{T1}{phv}{m}{n}
	\fontsize{10}{12}
	\selectfont
	\caption[Climatic variable selection]{Overview of process for identifying putative climatic selective drivers underlying latitudinal cline. We looked for climate variables that explained both the latitude of 356 \textit{M. cardinalis} occurences (`Climate-Latitude variables') and with traits (`Climate-Trait variables'). (A) For Climate-Latitude variables we extracted climate data from the specific location where a population was collected ($x_{i,0}$) throughout California and Oregon, USA (shown in map). For Climate-Trait variables, we extracted climatic data in two ways, but only for the 16 focal populations. (B) For point-estimated Climate-Trait variables, we extracted climate only from where a population was collected ($x_{i,0}$). (C) For spatially averaged Climate-Trait variables, we averaged over 1000 points in a 62-km radius climatic neighborhood ($x_{i,1}, x_{i,2}, \dots$), but only along stream habitats as \textit{M. cardinalis} is riparian.}
	\label{fig:Fig_VarSelect}
\end{figure}
